---
title: "HW_proteomics"
author: "Daria Ganaeva"
date: "20 04 2024"
output:
  
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Upload libraries

```{r}
library(limma)
library(ape)
library(dendextend)
library(RColorBrewer)
library(pvclust)
library(gplots)
library(NMF)
library(vegan)
library(Biobase)
library(DT)
library(ggplot2)
library(impute)
library(ggrepel)
library(clusterProfiler)
library(org.Hs.eg.db)
```

Upload the data

```{r}
data <- read.csv("peaks_data.csv")
```


View and check for NAs visually

```{r}
#View(data)

```

We have no empty samples, but there are missing (NA) values

Let's check if we have any duplicate genes and genes with half and more missing values and filter them out

```{r}

data_d <- data[which(!duplicated(data$Gene_id)), ]
rownames(data_d) <- data_d$Gene_id

col_number <- length(colnames(data_d[, -c(1:3)]))

nas <- rowSums(is.na(data_d[, names(data_d) %in% colnames(data_d[, -c(1:3)])]))

with_NA <- rownames(data_d)[nas > (col_number/2)]

data_f <- data_d[!rownames(data_d) %in% with_NA,]

nrow(data)
nrow(data_d)
nrow(data_f) 

```

We kept 2296 out of 3634 genes based on presence of duplicate genes and NAs in samples


Now we can proceed to KNN imputation:

```{r}
data_count <- as.matrix(data_f[, names(data_f) %in% colnames(data_f[, -c(1:3)])])
data_count <- t(data_count)
knn <- impute.knn(data_count, k = 5)
data_knn <- knn$data
data_knn <- t(as.matrix(data_knn))
head(data_knn, 2)
```


Check the distribution of the data

```{r}
pal <- brewer.pal(n = 9, name = "Set1")
data_experiment <- as.factor(c("control", "control", "control", "control", "control", "control", "control", "control", "control", "control", "control", "control", "control", "control", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment", "experiment"))

cols <- pal[data_experiment]
boxplot(data_knn, outline = FALSE, main = "Initial data", col = cols)
legend("topright", levels(data_experiment), fill = pal, bty = "n", xpd = T)


```
log transformation + normalization

```{r}
data_log <- log2(data_knn + 0.1)
data_norm <- normalizeQuantiles(as.matrix(data_log))
boxplot(data_norm, outline = FALSE, main = "Normalized data", col = cols)
legend("topright", levels(data_experiment), fill = pal, bty = "n", xpd = T)
```

Search for batch effect

```{r}
data_pca <- t(data_norm)
terapod_pca <- rda(data_pca, scale = TRUE)

rownames(data_pca) <- data_experiment

data_scores <- data.frame(data_pca,
                        scores(terapod_pca, display = "sites", choices = c(1, 2, 3), scaling = "sites"))

p_scores <- ggplot(data_scores, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = rownames(data_pca)), alpha = 0.5) +
  coord_equal(xlim = c(-3, 3), ylim = c(-3, 3)) + ggtitle(label = "Ordination of Principal Components") + theme_bw()
p_scores
```
We can see that our samples are clustered together based on their group status.
We can conclude that no batch-effect is captured by this plot. We can continue the analysis


let's build MA-plot to visually estimate the expression

```{r}
maplot <- function(X1, X2, pch = 21, main = "MA-plot", xlab = "Average log-expression", ylab = "Expression log-ratio", lpars = list(col = "blue", lwd = 2), ...){
  X <- (rowMeans(X2) + rowMeans(X1)) / 2
  Y <- rowMeans(X2) - rowMeans(X1)
  scatter.smooth(x = X, y = Y, main = main, pch = pch, xlab = xlab, ylab = ylab, lpars = lpars, ...)
  abline(h = c(-1, 0, 1), lty = c(2, 1, 2))
}

```

```{r}
maplot(data_norm[,1:14], data_norm[,15:33])
```

DE analysis

```{r}
# Expression data
expr_data <- as.matrix(data_norm)

# Sample data
pheno_data <- data.frame(data_experiment)
rownames(pheno_data) <- colnames(data_norm)
pheno_metadata <- data.frame(
  labelDescription = c("Experimental condition"),
  row.names=c("Condition"))
pheno_data <- new("AnnotatedDataFrame",
                 data = pheno_data,
                 varMetadata = pheno_metadata)

# Protein data
feature_data <- data.frame(Prot = rownames(expr_data))
rownames(feature_data) <- rownames(expr_data)
feature_metadata <- data.frame(
  labelDescription = c("Protein name"),
  row.names = c("Protein"))
f_data <- new("AnnotatedDataFrame",
              data = feature_data,
              varMetadata = feature_metadata)

# Merge together
exp_set <-
  ExpressionSet(assayData = expr_data,
                phenoData = pheno_data,
                featureData = f_data)
```

```{r}
X <- model.matrix(~ data_experiment, pData(exp_set))
fit <- lmFit(exp_set, design = X, method = "robust", maxit = 1000)
efit <- eBayes(fit)
```


```{r}
MA_limma <- function(efit, coef, n = 10, signif = TRUE, fdr = 0.05, lfc = 0, text = TRUE, cex.text = 0.8, col.text = "grey20", main = "MA-plot", xlab = "Average log-expression", ylab = "Expression log-ratio", pch = 19, pch.signif = 21, col = "darkgreen", alpha = 0.3, cex = 0.3, ...){
  
  R <- efit$coefficients[, coef]
  I <- efit$Amean
  
  col_btransp <- adjustcolor(col, alpha.f = alpha)
  
  plot(I, R, cex = cex, main = main, pch = pch, xlab = xlab, ylab = ylab, col = col_btransp, ...)
  abline(h = 0)
  
  if(signif){
    sign <- p.adjust(efit$p.value[, coef], method = "BH") <= fdr
    large <- abs(efit$coefficients[, coef]) >= lfc
    points(I[sign & large], R[sign & large], cex = cex*2, col = "orange2", pch = pch.signif)
  }
  
  if(text){
    ord <- order(efit$lods[, coef], decreasing = TRUE)
    top_n <- ord[1:n]
    text(I[top_n], R[top_n], labels = efit$genes[top_n, ], pos = 4, cex = cex.text, col = col.text)
  }
}


MA_limma(efit, coef = 2, n = 30)
```
Collect DE proteins

```{r}
my_list <- topTable(efit, coef = 2, n = 100)
dif_exp_set <- exp_set[fData(exp_set)$Prot %in% my_list$Prot, ]
```

```{r fig.height=10, fig.width=15}
dat <- as.matrix(exprs(dif_exp_set))

pal_blue_red <- colorpanel(75, low = "steelblue", mid = "black", high = "red")
heatmap.2(dat, col = pal_blue_red, scale = "row", key = TRUE, symkey = FALSE, density.info = "none", trace = "none", cexRow = 0.9, cexCol = 1, margins = c(4, 3), keysize = 0.8, key.par = list(mar = c(3, 0.1, 3, 0.1)))
```

```{r}
topTable(efit, coef = 2)
numGenes <- nrow(exprs(exp_set))
full_list <- topTable(efit, number = numGenes)
full_list <- full_list[full_list$adj.P.Val <= 0.05,]
write.csv(full_list, "DE_full_proteomics.csv")

nrow(full_list)
```

Volcano plot

```{r}
volcano_list <- full_list

volcano1 <- ggplot(data = volcano_list, aes(x = data_experiment, y = -log10(adj.P.Val))) + geom_point()

volcano2 <- volcano1 + geom_vline(xintercept=c(-0.6, 0.6), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red")

volcano_list$diffexpressed <- "NO" 

volcano_list$diffexpressed[volcano_list$logFC > 2 & volcano_list$adj.P.Val < 0.05] <- "UP"

volcano_list$diffexpressed[volcano_list$logFC < -2 & volcano_list$adj.P.Val< 0.05] <- "DOWN"

volcanodif1 <- ggplot(data = volcano_list, aes(x = logFC, y = -log10(adj.P.Val), col = diffexpressed)) + 
  geom_point() + 
  theme_minimal() +
      geom_text_repel(data = volcano_list[1:10, ], aes(label = volcano_list[1:10, ]$Prot))


volcanodif1
```

Count number of DE proteins:

```{r}
length(which(volcano_list$logFC > 2 & volcano_list$adj.P.Val < 0.05))
length(which(volcano_list$logFC < -2 & volcano_list$adj.P.Val < 0.05))
```
GO Enrichment analysis

```{r}

data_pv <- full_list[full_list$adj.P.Val < 0.05,]

data_pv_down <- data_pv[data_pv$logFC < -2,]

data_pv_up <- data_pv[data_pv$logFC > 2,]

write.csv(data_pv_down, "Prot_down.csv")
write.csv(data_pv_up, "Prot_up.csv")

```

GO enrichment analysis was performed using web-service https://biit.cs.ut.ee/gprofiler/.


Conclusion:

GO terms regarding oxydoreductase activity, electron transport, catabolic processes, and mitochondrial structures (e.g. inner and outer membranes) are enriched among up-regulated proteins, which possibly indicate the activation of cellular respiration and oxidative metabolism. Hence, enriched antioxidant activity may be a possible mechanism to prevent the associated increased production of reactive oxygen species. Enriched terms associated with lysosomes may be a mark of high degree of protein and lipids damage. This overall indicates that the cells experience oxydative stress and associated events leading to disruption of cell functions possibly leading to macro-scale events such as inflammation.

On the other hand, we observe processes and structures associated with cytoskeleton (e.g. actin filaments, intermediate filaments) among down-regulated terms. Together with down-regulated cell junction terms, it may possibly indicate that cells undergo serious shape and morphological changes. Among down-regulated proteins L-serine transport and metabolism enrichment terms were observed. We are not able to make a firm conclusion about the event, but it can be linked with lactic acid and nitrogen metabolism, which are key processes in heart cells.

Various GO terms associated with cell and protein localization and exosomes were among both up- and down-regulated proteins, they are very broad terms, which makes it hard to interprete these observations.
